
\chapter*{Introduction}

The vertical distribution of fish eggs in a water column is determined
by the buoyant forcing and turbulent mixing.  A model of the steady
state distribution has been developed by Sundby \shortcite{sund83}.
Westg{\aa}rd \shortcite{west89} developed two numerical models for the time
evolution of the concentration. These models are written in FORTRAN
and use the GPGS library for graphics.

The commercial software system Matlab and the free alternative Octave
provide interactive command-line environments for numerical computations
and visualisation.  VertEgg is a toolbox for scientific work on
vertical egg distributions in these environments. It contains tools for
analysing observed distributions and performing numerical simulations
including pre- and post-processing of the results. VertEgg is
\emph{not} a finished application, the tools are intended to be used
interactively or sewed together by the user to make the application
that solves her problem. This approach should give the competent user a
powerful and flexible environment for research on vertical distributions.

The author believes strongly in learning by problem solving and
examples. Therefore the main part of the manual is the example chapter.
The examples ranges from simple illustrations of the functions in the
toolbox, by verification and sensitivity studies to a ready-to-run
egg-distribution model complete with file I/O.
These examples are also available online as \emph{scripts}, programs in
the internal programming language of the packages. The recommended way
to make small applications is to take the closest example script and
try out modifications.

The first two chapters give an introduction to the theory of vertical
egg distributions and the numerical method used. The manual also has a
chapter containing the reference manual, a systematic description of the
tools provided in VertEgg.

% Ta noe om tilgjengelighet, copyright ??

